Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-03-19 Reviews-response.tex

%!LW recipe=pdflatex
\documentclass{article}

\usepackage{tikz-cd}
\usepackage{xargs}
\usepackage{amsmath}

\input{preamble}
\input{macros}
\input{sample.tikzstyles}
\input{hypergraph.tikzstyles}
\input{hypergraph.tikzdefs}

\newcommand\id{\textsf{id}}
\newcommand\sym{\textsf{sym}}

\begin{document}

\title{LICS24 Reviews Response}
\author{}

\maketitle

\section*{Review 1}
\paragraph{About theorem 6.5 (the central result of the paper: full completeness)}
\textit{I am doubtful about the statement because it seems to me that $[sym] = [id]$,
hence the functor $[-]$ is not faithful.
Here is a sketch of the argument: $[\sym]$ is a e-hypergraph with two vertices
where the inputs and the outputs are inverted. Now we embed it into a
hierarchical edge by applying the bottom left equality of Figure 3.
We get a cospan with 2 external inputs, 2 external ouputs, 2 strict internal
inputs, 2 strict internal outputs (the sets of strict internal inputs and
outputs are the same, but they are labelled differently by $f_{int}$ and $g_{int}$). We
do the same thing with $[\id]$ and notice that the two resulting cospans are
isomorphic. Thus they are equal in the category where cospans are quotiented by
isos.}

We assume that in $[\sym] = [\id]$ it was meant that $[\sym] = [\id \otimes \id]$, otherwise the equality does not typecheck.

By the definition $[\sym]$ is 
\[
    n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
\]
represented as
\begin{figure}[h!]
    \[
    \scalebox{0.5}{
        \tikzfig{reviews/sym}
    }
    \]
\end{figure}

And $[\id \otimes \id]$ is \[
    n \xrightarrow{f'_{ext}} n'' \xrightarrow{f'_{int}} \mathcal{G} \xleftarrow{g'_{int}} m'' \xleftarrow{g'_{ext}} m
\]
represented as 
\begin{figure}[h!]
    \[
    \scalebox{0.5}{
        \tikzfig{reviews/id}
    }
    \]
\end{figure}

Clearly, it should be the case that $\{u_1 \mapsto u_1, \ldots u_4 \mapsto u_4\}$ and similarly for $w_i$. I think this does not follow from our definition of isomorphic cospans, so perhaps we should've required that $\alpha$ and $\gamma$ are identities (below).
\[
    \scalebox{0.75}{
        \tikzfig{combinatorial_semantics/isomorphic_e_cospans}
    }
\]
Then, the cospans above are not isomorphic. Consider the cases.
\begin{enumerate}
    \item ${v_1 \mapsto v_1}$ and ${v_2 \mapsto v_2}$ under $\beta$. Then $g_{int};\beta(w_4) \not = g'_{int}(w_4)$
    \item ${v_1 \mapsto v_2}$ and ${v_2 \mapsto v_1}$ under $\beta$. Then $f_{int};\beta(w_4) \not = f'_{int}(w_4)$
\end{enumerate}
This means that the corresponding diagram does not commute and therefore the cospans are not isomorphic.

\paragraph{In the proof}
In the first case, it is immediate that $[f] \not = [g]$, since equal terms
have weak normal forms with the same number of components.
\textit{This argument does not seem to be valid in presence of equations.
Consider an equation $a = a + b$ for example, where a and b are constants of the
signature. Then}
\[
    [a] = [a + b] = [a] + [b]
\]

We only consider equations of SMC terms, \textit{i.e.}, with no $+$ on either side of $=$.

\end{document}

### 2024-03-19 Reviews-response.tex

Clearly, it should be the case that $\{u_1 \mapsto u_1, \ldots u_4 \mapsto u_4\}$ and similarly for $w_i$. I think this does not follow from our definition of isomorphic cospans, so perhaps we should've required that $\alpha$ and $\gamma$ are identities (below).
\[
    \scalebox{0.75}{
        \tikzfig{combinatorial_semantics/isomorphic_e_cospans}
    }
\]
Then, the cospans above are not isomorphic. Consider the cases.
\begin{enumerate}
    \item ${v_1 \mapsto v_1}$ and ${v_2 \mapsto v_2}$ under $\beta$. Then $g_{int};\beta(w_4) \not = g'_{int}(w_4)$
    \item ${v_1 \mapsto v_2}$ and ${v_2 \mapsto v_1}$ under $\beta$. Then $f_{int};\beta(w_4) \not = f'_{int}(w_4)$
\end{enumerate}
This means that the corresponding diagram does not commute and therefore the cospans are not isomorphic.

\paragraph{In the proof}
In the first case, it is immediate that $[f] \not = [g]$, since equal terms
have weak normal forms with the same number of components.
\textit{This argument does not seem to be valid in presence of equations.
Consider an equation $a = a + b$ for example, where a and b are constants of the
signature. Then}
\[
    [a] = [a + b] = [a] + [b]
\]

We only consider equations of SMC terms, \textit{i.e.}, with no $+$ on either side of $=$.

\paragraph{About Proposition A.14 (existence of the pushout complement)}
\textit{Why doesn't the pushout complement trivially exist since you already assume
given a pushout square?}

It is indeed the case.
The formulation should've been consider a morhisms $\mathcal{K} \to \mathcal{L} \to \mathcal{G}$.

\paragraph{In the proof:}
Then, no-dangling and no-identification
conditions are necessary for pushout complement to exist [26].
\textit{Are those conditions sufficient? Can you give a more precise reference? It seems
to me that [26] (in particular Proposition 9 about the existence of pushout
complements) focuses on the category of graphs only, so I don't see how it
applies to the category of e-hypergraphs.}

I do not know if those conditions are sufficient. 
I am only sure the three conditions of A.14 are sufficient.
We only proved that the pushout exists for a certain class of morphisms hence the additional conditions.
We did not, however, prove that the pushout does not exist for other classes of morphisms.
The no-dangling condition ensures that the sources or targets of the edge which is not a part of a match do not get deleted.
If it was the case then there would be no way to obtain the original graph by computing a pushout as there would be not enough vertices.
Similarly with no-identification condition, if $m$ was to identify two vertices of $\mathcal{L}$ while these were not identified by $f$ then there would be no way of getting $\mathcal{G}$ by only knowing $\mathcal{L}$, $\mathcal{L^{\bot}}$ and $\mathcal{K}$ as computing the union of $\mathcal{L}$ and $\mathcal{L^{\bot}}$ would produce more vertices than there was in $\mathcal{G}$.
The proposition ensures that after we remove the image of $\mathcal{L}$ from $\mathcal{G}$ (excluding the image of $f;m$), all the morphisms in the pushout square are suitable for our computation of the pushout.
There is a claim about the necessity and sufficiency of no-identification and no-dangling conditions for the pushout to exist in $\mathbf{Hyp}(\Sigma)$ in Proposition 3.17 of [3].

\paragraph{About Proposition A.15 (uniqueness of the boundary complement)}
The pushout in $\mathbf{EHyp}(\Sigma)$ should also be a pushout on the
underlying sets of vertices and hyperedges.
\textit{Why is that? Is it always the case for any pushout in $\mathbf{EHyp}(\Sigma)$?}

This is by construction as we compute the disjoint union of vertices and edges and identify them along the corresponding morphisms.
This construction yields the pushouts on the sets of vertices and edges.
The construction of the corresponding consistency and child relations are orthogonal as these do not change the sets of edges and vertices.
This is at least the case for a particular set of morphisms that we consider.

\textit{Then, since all the arrows are monos}
\textit{What if some vertex is both input and output in $\mathcal{L}$? Then $i' + j' \to L$ is not a
monomorphism, is it?}

Yes, then $i' + j' \to L$ is not a monomorphism.

\paragraph{About remark 2 (restriction to signatures without any symbol of arity $0 \to 0$)}
\textit{By incident vertices, do you mean input and output vertices? I think there is
still a problem if you consider an hypergraph $L$ composing an edge of arity $0 \to
1$ with an edge of arity $1 \to 0$: the boundary conditions prevent finding a match
of $L$ in any $G$.}

Yes, this is correct.

\paragraph{About Def 4.1 (e-hypergraphs)}
\textit{Why don't you require the condition (3) for t(e) as well?}

That's a typo. It should've been similar to consistency relation.

\textit{The consistency equivalence relation must additionally satisfy that if $v \in s(e)$ or $v \in t(e)$ then $e \consistency v$}
\textit{What if 'e' is a top-level edge?}

Should've been if $e$ is not a top-level edge.

\paragraph{About def 4.2 (e-hypergraph homomorphism)}

\textit{Condition (2): what if the image of 'v' or 'e' is top-level?}

Then we do not require such an equality, \textit{i.e.}, it might hold or might not meaning that a top-level vertex of edge might end up either top-level or nested under a homomorphism but nested edges and vertices should preserve their predecessors.

\paragraph{About def 4.5 (well-typed MDA hypergraphs)}
\textit{Why aren't non top-level hierarchical edges required to be well-typed?}

All hierarchical edges must be well-typed. The `apex' refers to the apex of a cospan, \textit{aka}, the carrier.

\paragraph{About def 5.2 (extended boundary complement)}
\textit{Isn't requiring that $i+j \to L^\bot$ is a mono too strong? For example if you
want to account for an equality relating some term involving the identity
morphism (for which all nodes are both inputs and outputs).}

Yes, it is strong and was needed to make the proofs easier. 
Dropping this condition is a future work.

\textit{Are monomorphisms in the category of hypergraphs the morphisms that are
injective on edges and vertices?}

Yes.

\paragraph{Minor remarks / typos}

\textit{In the figure above, shouldn't there be $n$ (rather than just 1) pairs of black
dots in each dashed subbox? (p2)}

There is a dot for each unused input of the corresponding instance within a box.
For example, $*$ requires $a$ and $2$ and therefore the wires from another copy of $2$ and from $1$ are terminated with dots.

\textit{removing from this occurrence everything that has no pre-image in $\mathcal{K}$ (p6)}
\textit{Isn't $C$ rather constructed by removing from $G$ everything from $L$ except for what
comes from $K$?}

Yes, this formulation is equivalent. We are removing $m(L)$ from $G$ with the exception of nodes and edges that have a pre-image in $K$, \textit{i.e.}, those that come from $K$.
\textit{What is a monogamous cospan? p(7)}

Apparently the definition of a monogamous cospan was overlooked.

\textit{This is confusing: you are not defining any equivalence relation here. Rather, 
this family of equivalence relations is provided by each hypergraph (if I
understand correctly). Def 4.1 (p7)}

Yes, it is indeed a family of relations.

\textit{Shouldn't a homomorphism preserves labelling?}

It preserves labelling as per the condition (1) on such a morphism: it should be a hypergraph-homomorphism, which, by definition, preserves labelling.

\textit{Unfolding definitions, it follows that}
$f = g \implies \llbracket f \rrbracket \Rrightarrow^{*} \llbracket g \rrbracket$.

\textit{The literal meaning makes this statement trivial: if $f = g$, then $[f]$ rewrites to
$[g]$ in 0 steps.
If we replace $f = g$ with $p_i(f) = p_i(g)$, where $f$ and $g$ are unquotiented terms and
$p_i$ is the canonical projection, this is wrong, because it would entail that the
rewriting relation is symmetric, which is not the case, unless we consider} $\Rrightarrow^{*}$
\textit{that each equation generates two rewrite rules, one for each direction of the
equality (if is the case, this should be said explicitly). Corollary 6.6 suffers
from the same defect, and the proof of Theorem 6.5 as well.}

$\llbracket \mathcal{E} \rrbracket$ consists of $\langle l, r \rangle$ and $\langle l, r \rangle$ for each $l = r$ in $\mathcal{E}$.

\textit{Aren't they partial functions, defined on any non-top level elements only?}
Yes

\textit{Condition (5) implicitly assumes that f(v) is not top-level, because the
consistency relation is only defined for such elements.}

It should be read as either all $f(v_i)$ are pair-wise consistent, or consistency for them is undefined, \textit{i.e.}, all of them are top-level.

\textit{In which category is this pushout proven to exist? Clearly not in the category
$\mathbf{EHyp}(\Sigma)$ of hypergraphs, since just after the proof you write that it was not
proven to be a hypergraph.}

It becomes a pushout in $\mathbf{EHyp}(\Sigma)$ right after it is proved that it is a valid e-hypergraph (Proposition A.12).
Probably the proof should've been ordered the other way around.

\textit{In the proof, you consider the identity relation on edges, then its
reflexive-transitive-symmetric closure. Doesn't it coincide with the identity
relation? Furthermore, doesn't quotienting a set X by the identity relation
yields the same set X?
Do you really need to take the transitive closure of the relation on vertices,
since f and g are monomorphisms?}

This was just to make the presentation uniform.
\end{document}

### 2024-03-20 Reviews-response.tex

\textit{In which category is this pushout proven to exist? Clearly not in the category
$\mathbf{EHyp}(\Sigma)$ of hypergraphs, since just after the proof you write that it was not
proven to be a hypergraph.}

It becomes a pushout in $\mathbf{EHyp}(\Sigma)$ right after it is proved that it is a valid e-hypergraph (Proposition A.12).
Probably the proof should've been ordered the other way around.

\textit{In the proof, you consider the identity relation on edges, then its
reflexive-transitive-symmetric closure. Doesn't it coincide with the identity
relation? Furthermore, doesn't quotienting a set X by the identity relation
yields the same set X?
Do you really need to take the transitive closure of the relation on vertices,
since f and g are monomorphisms?}

This was just to make the presentation uniform.

\section*{Review 2}
\textit{The discussion about Figure 1 in Section (1.1) does not provide enough details to understand what the graph describes. In particular, in the e-hypergraph formalism, I am not sure I understand what the black nodes inside the equivalence classes mean. For instance, why are these nodes only connected to the bottom of the boxes and never to the top?}

Black dots correspond to delete morphisms of the Cartesian category, \textit{i.e.}, they discard the unused input and hence are only conected to the bottom as inputs flow from bottom to top.

\textit{Is the free monoid monad the list monad?}

Yes

\textit{Without further assumptions (e.g., some form of adhesivity), pushout complements may not be unique, assuming it exists. Is there any guarantee in the DPOI framework for hypergraphs with interfaces? How does the construction of boundary complement (Definition 3.8) relate to initial pushouts (see, for instance, Chapter 6 of Fundamentals of Algebraic Graph Transformation Ehrig et a. 2006, page 125)?}

The category of plain hypergraphs ($\mathbf{Hyp}$), as used in [3], is adhesive.
We do not know if the category of e-hypergraphs ($\mathbf{EHyp}$) has any adhesive properties, however, in this work we are only interested in pushouts along monos and monomorphic matches and for such morphisms the pushout complement is unique as shown in A.15.
The boundary complement definition is a generalisation of a similar concept from [4] which is needed to make sure we remain within needed class of morphisms when computing the complement and the pushout.
We can not immediately relate initial pushouts to the boundary complement.

\textit{One could argue that since the consistency relation is built from the child relation, it should not be part of the tuple defining the e-hypergraph.}

?

\textit{In Figure 5, why do $a_1$, ... $a_n$, and $a'_1$, ..., $a'_k$ appear as internal interfaces on the rightmost hypergraph?}

Internal interface is external interface plus strict internal interface. Since both $a_i$'s and $a'_j$'s are external interfaces they should be present in internal interface as well.

\section*{Review 3}
\textit{However, I think the authors could improve the discussion of the 0-ary case and be more convincing about the fact that this could be handled without much difficulty. Otherwise, this would dramatically restrict the domain of application of those results.}

One workaround is to add more context to the left-hand side of a rewrite rule. 
Assume we have a rewrite rule $f \Rrightarrow g \otimes a$ where $a$ is a scalar.
If the occurrence of $f$ within $\mathcal{G}$ is top-level, then we can proceed as usual.
If, on the other hand, the occurrence is nested, then there is no way to specify that $a$ should have the same parent as $f$ had.
The workaround can be seen in the following example.

\[
    \scalebox{0.5}{
        \tikzfig{reviews/handling_scalars}
    }
\]

$\mathcal{F}, \mathcal{G}_i$ are variable-labelled e-hypergraphs that can be unified with concrete e-hypergraphs.
For example, $\mathcal{F}$ corresponds to $h$ and $\mathcal{G}_1$ corresponds to $k$.

\section*{Review 4}

Review 4 had no questions.

\end{document}

### 2024-03-21 Reviews-response.tex

We assume that in $[\sym] = [\id]$ it was meant that $[\sym] = [\id \otimes \id]$, otherwise the equality does not typecheck.

By the definition $[\sym]$ is 
\[
    n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
\]
represented as
\begin{figure}[h!]
    \[
    \scalebox{0.5}{
        \tikzfig{reviews/sym}
    }
    \]
\end{figure}

And $[\id \otimes \id]$ is \[
    n \xrightarrow{f'_{ext}} n'' \xrightarrow{f'_{int}} \mathcal{G} \xleftarrow{g'_{int}} m'' \xleftarrow{g'_{ext}} m
\]
represented as 
\begin{figure}[h!]
    \[
    \scalebox{0.5}{
        \tikzfig{reviews/id}
    }
    \]
\end{figure}

Clearly, it should be the case that $\{u_1 \mapsto u_1, \ldots u_4 \mapsto u_4\}$ and similarly for $w_i$. I think this does not follow from our definition of isomorphic cospans, so perhaps we should've required that $\alpha$ and $\gamma$ are identities (below).
However, the interfaces are ordered according to a remark on page 21 in [3], so the only possible isomorphism between $n'$ and $n''$ (respectively, $m'$ and $m''$) is identity.
\[
    \scalebox{0.75}{
        \tikzfig{combinatorial_semantics/isomorphic_e_cospans}
    }
\]
Then, the cospans above are not isomorphic. Consider the cases.
\begin{enumerate}
    \item ${v_1 \mapsto v_1}$ and ${v_2 \mapsto v_2}$ under $\beta$. Then $g_{int};\beta(w_4) \not = g'_{int}(w_4)$
    \item ${v_1 \mapsto v_2}$ and ${v_2 \mapsto v_1}$ under $\beta$. Then $f_{int};\beta(w_4) \not = f'_{int}(w_4)$
\end{enumerate}
This means that the corresponding diagram does not commute and therefore the cospans are not isomorphic.

@@

\textit{One could argue that since the consistency relation is built from the child relation, it should not be part of the tuple defining the e-hypergraph.}

?

\textit{In Figure 5, why do $a_1$, ... $a_n$, and $a'_1$, ..., $a'_k$ appear as internal interfaces on the rightmost hypergraph?}

Internal interface is external interface plus strict internal interface. Since both $a_i$'s and $a'_j$'s are external interfaces they should be present in internal interface as well.
The figure is essentially saying that we sum both internal interfaces and add extra copy of external interface (to account for the inputs and outputs of this new hierarchical edge).

### 2024-04-17 EGraphsExtendedAbstract.tex
the submitted version of extended abstract

\author{Anonymous}

@@

This work in progress aims to propose a generalisation of e-graphs for arbitrary SMTs which is build by enriching the underlying categorical structure of a given SMT.
Then, it can be seen that original e-graphs is the case of our construction applied to Cartesian SMTs.
Apart from building the foundation for the application of e-graphs to a richer class of theories, this work also provides a new perspective on solving the notorious issues of alpha-equivalence and binding that arise when, for example, one wants to build e-graphs for functional languages~\cite{koehler2022sketchguided}.

@@

To express the notion of e-classes of SMT terms we consider an SMT enriched over a category of semilattices as defined below.

@@

\section{E-graphs as terms of a Cartesian SMT}
We are now ready to present how (asyclic) e-graphs can be considered as terms of a Cartesian SMT.
Let $\Sigma$ be a monoidal signature in which every generator $c$ has a single output and $\mathcal{E}$ be the set of equations on $\Sigma$-terms. 
Then let $\textbf{CSMT}(\Sigma_{C},\mathcal{E}_{C})$ be the Cartesian SMT as introduced in Example~\ref{example:csmt}.
\begin{proposition}
    Acyclic e-graphs over a set of generators $\Sigma$ and equivalence classes of terms on $\textbf{CSMT}^{+}(\Sigma_{C},\mathcal{E}_{C})$ are in one-to-one correspondence.
\end{proposition}

@@

Nodes are encapsulated in dashed boxes indicating equivalence classes of nodes, or, more generally, of subgraphs. 
Initially each equivalence class has a single node. 
The first rewrite rule, replacing multiplication by the more efficient \emph{shift-left} operator creates the first non-unitary equivalence class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) operator nodes, with a new sharing, that of node $a$. 
This second e-graph illustrates the other difference, besides sharing, between conventional e-graphs and e-string diagrams, namely the fact that in the former edges connect directly to nodes inside the equivalence class, whereas in the latter edges (wires) connect to the equivalence class itself. 
Explicit discarding ($!$-generator) nodes indicate which of the class-level edges are connected to which nodes inside the class. 
Generally, the application of rewrite rules to e-graphs correspond to a sequence of equations applied to e-string diagrams.
A step-by-step application of such equations for the rewrite rule from $(a)$ to $(b)$ and $(b)$ to $(c)$ can be seen in Figure~\ref{fig:e-graph-example-a-b} and Figure~\ref{fig:e-graph-example-a-b} respectively.
The equations applied are the ones from Figure~\ref{fig:string-equations} and naturality of delete and copy.
One can note that the size of the eventual e-string diagram is comparable to the size of the corresponding e-graph.
\end{example}

We can also translate an arbitrary e-graph into an e-string diagram as shown belownoting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding in each component all unnecessary inputs.
\[
    \tikzfig{categorical-semantics/egraph-translation-1}
\]
Note, further, that in the Cartesian case we can restrict to generating morphisms $c$ to have a single output,  without loss of generality,  by applying the universal properties of the Cartesian product.  This means the (informal) translation given above is fully general for acyclic e-graphs.
We believe that e-graphs with cycles can be represented as terms in Traced Cartesian SMTs and accounting for a trace above seems to be routine.

While we did not explicitly define how exactly equations are applied to e-string diagrams resorting rather to intuition, e-string diagrams can be formalised as concrete combinatorial structures by following an established approach of converting string diagrams for non-enriched SMCs to hypergraphs~\cite{bonchi_string_2022,bonchi_string_2022-1,bonchi_string_2022-2}.
We denote the corresponding combinatorial structure an e-hypergraph and claim that applying equations to e-string diagrams can be mimicked by performing a \textit{double pushout} rewriting on e-hypergraphs.
This allows us to express a complex e-graph rewriting, that includes creation of e-classes, merging, and computing congruence closure as a sequence of e-hypergraph rewriting steps which as we believe could facilitate reasoning about e-graphs.
The difference between e-string diagrams and e-hypergraphs is that the former are implicitly quotiented by the equations in Figure~\ref{fig:string-equations} while for the latter these equations become structural rewrite rules.
One can notice that we unfolded e-string diagrams in Figure~\ref{fig:e-graph-example-b-c} to identify a redex and then to factor out the nodes to achieve a better sharing.
This would also be the case when applying rewrite rule for e-hypergraph representation in a na\"ive way.
The work towards finding redexes and achieving maximal sharing without excessive unfolding is in progress.

@@

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \scalebox{0.5}{
        \tikzfig{categorical-semantics/e-string-substitution}
        }
    \]
    \caption{E-string diagrammatic substitution example.}
    \label{fig:e-string-substitution}
\end{figure*}

\section{E-string diagrams for lambda calculus}
In this section we will show how a particular kind of SMTs can provide a new perspective on solving the long-standing issues of $\alpha$-equivalence and binding when using e-graphs for languages like lambda calculus.
The first issue appears due to the fact that lambda calculus terms are quotiented by renaming of bound variables.
If such renaming is not accounted for by the language encoding, \textit{e.g.,} by using \textit{De Brujin} indices, populating an e-graph with $\alpha$-equivalent terms increases the size of an e-graph without adding any meaningful equivalences.
The other issue is that having binding entails having a substitution and computing substitution in e-graphs is hard as it is not a mere substitution of terms but rather a subsittution of eqivalence classes that represent a lot of terms at once.
An example of a substitution can be seen in Figure~\ref{fig:e-graph-substitution}.
Notably, the terms that correspond to explicit substitution can be discarded at the end.

By recalling the correspondence between Cartesian Closed SMCs and lambda-calculus we can consider the corresponding semilattice enriched Cartesian Closed SMTs.
Apart from the generators in $\Sigma_{C}$ such theories also contain \textit{application} and $\textit{abstraction}$ maps that correspond to $app$ and $\lambda$ respectively.
A complete description of these maps and the corresponding equations that they obey can be found in~\cite{ghica-zanassi2023string}.
The corresponding e-string diagram can be seen in Figure~\ref{fig:e-string-substitution}.
The half-circle nodes correspond to application maps and the rounded rectangle with a wire attached to its left side is the abstraction map where the attached wire corresponds to the argument of the abstraction.
A distinguishable feature of these string diagrams is that they are automatically quotiented by $\alpha$-equivalence and that a substitution is merely a rewrite step that glues an argument of the application to the attached wire of the abstraction.

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \scalebox{0.5}{
        \tikzfig{categorical-semantics/egraph-translation-step-by-step-b-c}
        }
    \]
    \caption{Example translation from $(b)$ to $(c)$.}
    \label{fig:e-graph-example-b-c}
\end{figure*}

>>>>

\author{Chris Barrett}
\affiliation{
  \institution{University of Oxford}
  \country{United Kingdom}
}

\author{Aleksei Tiurin}
\affiliation{
  \institution{University of Birmingham}
  \country{United Kingdom}
}

\author{Dan Ghica}
\affiliation{
  \institution{University of Birmingham}
  \country{United Kingdom}
}

@@

Our work in progress aims to propose a generalisation of e-graphs for arbitrary SMTs which is build by enriching the underlying categorical structure of a given SMT.
Then, it can be seen that original e-graphs is the case of our construction applied to Cartesian SMTs.
Apart from building the foundation for the application of e-graphs to a richer class of theories, this work also provides a new perspective on solving the notorious issues of alpha-equivalence and binding that arise when, for example, one wants to build e-graphs for functional languages~\cite{koehler2022sketchguided}.

@@

To express the notion of e-classes of SMT terms we consider an SMT which is represented by an SMC enriched over a category of semilattices as defined below.

@@

\section{E-graphs as terms of a Cartesian SMT}
We are now ready to present how (acyclic) e-graphs can be considered as terms of a Cartesian SMT.
Let $\Sigma$ be a monoidal signature in which every generator $c$ has a single output and $\mathcal{E}$ be the set of equations on $\Sigma$-terms. 
Then let $\textbf{CSMT}(\Sigma_{C},\mathcal{E}_{C})$ be the Cartesian SMT as introduced in Example~\ref{example:csmt}.
\begin{proposition}
    For each acyclic e-graph over a set of generators $\Sigma$ and a set of equations $\mathcal{E}$ that have at most one root e-class, there is a corresponding equivalence class of terms in $\textbf{CSMT}^{+}(\Sigma_{C},\mathcal{E}_{C})$.
    Moreover, if $e_1$ and $e_2$ are e-graphs such that $e_2$ is obtained from $e_1$ after applying some rewrite rule (\textit{i.e.}, $e_2$ is obtained after adding new terms into $e_1$ and performing the necessary merges and repairs), then $s_2$ can be obtained from $s_1$ by applying the equations from Figure~\ref{fig:string-equations} and $\mathcal{E}_{C}$ where $s_1$ and $s_2$ are the corresponding e-string diagrams for $e_1$ and $e_2$ respectively.
\end{proposition}

@@

Nodes are encapsulated in dashed boxes indicating equivalence classes of nodes, or, more generally, of subgraphs. 
Initially each equivalence class has a single node. 
The first rewrite rule, replacing multiplication by the more efficient \emph{shift-left} operator creates the first non-unitary equivalence class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) operator nodes, with a new sharing, that of node $a$. 
This second e-graph illustrates the other difference, besides sharing, between conventional e-graphs and e-string diagrams, namely the fact that in the former edges connect directly to nodes inside the equivalence class, whereas in the latter edges (wires) connect to the equivalence class itself. 
Explicit discarding ($!$-generator) nodes indicate which of the class-level edges are connected to which nodes inside the class. 
Generally, the application of rewrite rules to e-graphs correspond to a sequence of equations applied to e-string diagrams.
A step-by-step application of such equations for the rewrite rule from $(a)$ to $(b)$ and $(b)$ to $(c)$ can be seen in Figure~\ref{fig:e-graph-example-a-b} and Figure~\ref{fig:e-graph-example-a-b} respectively.
The equations applied are the ones from Figure~\ref{fig:string-equations} and naturality of delete ($!$) and copy ($\Delta$).
One can note that the size of the eventual e-string diagram is comparable to the size of the corresponding e-graph.
\end{example}

We can then translate an arbitrary e-graph into an e-string diagram as shown below, noting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding in each component all unnecessary inputs.
\[
    \tikzfig{categorical-semantics/egraph-translation-1}
\]
Note, further, that in the Cartesian case we can restrict to generating morphisms $c$ to have a single output,  without loss of generality,  by applying the universal properties of the Cartesian product.  This means the (informal) translation given above is fully general for acyclic e-graphs.
We believe that e-graphs with cycles can be represented as terms in Traced Cartesian SMTs and accounting for a trace above seems to be routine.

While we did not explicitly define how exactly equations are applied to e-string diagrams resorting rather to intuition, e-string diagrams can be formalised as concrete combinatorial structures by following an established approach of converting string diagrams for non-enriched SMCs to hypergraphs~\cite{bonchi_string_2022,bonchi_string_2022-1,bonchi_string_2022-2}.
We denote the corresponding combinatorial structure an e-hypergraph and claim that applying equations to e-string diagrams can be mimicked by performing a \textit{double pushout} rewriting of e-hypergraphs.
This allows us to express a complex e-graph rewriting, that includes creation of e-classes, merging, and computing congruence closure as a sequence of e-hypergraph rewriting steps which as we believe could facilitate reasoning about e-graphs.
The difference between e-string diagrams and e-hypergraphs is that the former are implicitly quotiented by the equations in Figure~\ref{fig:string-equations} while for the latter these equations become structural rewrite rules.
One can notice that we unfolded e-string diagrams in Figure~\ref{fig:e-graph-example-b-c} to identify a redex and then to factor out the nodes to achieve a better sharing.
This would also be the case when applying rewrite rules for e-hypergraph representation in a na\"ive way.
The work towards finding redexes and achieving maximal sharing without excessive unfolding is in progress.

@@

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \scalebox{0.5}{
        \tikzfig{categorical-semantics/e-string-substitution-2}
        }
    \]
    \caption{E-string diagrammatic substitution example.}
    \label{fig:e-string-substitution}
\end{figure*}

\begin{figure*}[t!]
    \vspace{-3cm}
    \centering
    \[
        \hspace{1.3cm}
        \resizebox{0.8\textwidth}{!}{
        \tikzfig{categorical-semantics/egraph-translation-step-by-step-b-c}
        }
    \]
    \caption{Example translation from $(b)$ to $(c)$.}
    \label{fig:e-graph-example-b-c}
\end{figure*}

\section{E-string diagrams for lambda calculus}
In this section we will show how a particular kind of SMTs can provide a new perspective on solving the long-standing issues of $\alpha$-equivalence and binding when using e-graphs for languages like lambda calculus.
The first issue appears due to the fact that lambda calculus terms are quotiented by renaming of bound variables.
If such renaming is not accounted for by the language encoding, \textit{e.g.,} by using \textit{de Brujin} indices, populating an e-graph with $\alpha$-equivalent terms increases the size of an e-graph without adding any meaningful equivalences.
The other issue is that having binding entails having a substitution and computing substitution in e-graphs is hard as it is not a mere substitution of terms but rather a substitution of equivalence classes that represent a lot of terms at once.
An example of a substitution can be seen in Figure~\ref{fig:e-graph-substitution}. An explicit substitution operator is the part of the syntax and the substitution of e-classes is defined inductively on the structure of all terms that these e-classes represent, \textit{i.e.}, computing a substitution of e-classes is achieved by computing a pair-wise substitution for all the instances of these e-classes.
Notably, the terms that correspond to explicit substitution can be discarded at the end.

By recalling the correspondence between Cartesian Closed SMCs and lambda-calculus we can consider the corresponding semilattice enriched Cartesian Closed SMTs.
Apart from the generators in $\Sigma_{C}$ such theories also contain \textit{application} and $\textit{abstraction}$ maps that correspond to $app$ and $\lambda$ respectively.
A complete description of these maps and the equations that they obey can be found in~\cite{ghica-zanassi2023string}.
The e-string diagram version of the previous example can be seen in Figure~\ref{fig:e-string-substitution}.
The half-circle nodes correspond to application maps and the rounded rectangles with a wire attached to their right sides are the abstraction maps where the attached wire corresponds to the argument of the abstraction.
A distinguishable feature of these string diagrams is that they are automatically quotiented by $\alpha$-equivalence and that a $\beta$-reduction is merely a rewrite step that glues an argument of the application to the attached wire of the abstraction and the substitution is just string diagram composition.
Another advantage of string diagrams is that every occurrence of the bound variable is essentially the same wire as opposed to de Brujin indices, so there is no need to modify any indices, the whole substitution is just a matter of reattaching the wires non-destructively.
In particular, note how the substitution within $y + y$ was completely avoided.
% The latter follows from the definition of the respective adjunction that defines these maps.
